Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2022-11-17 nakayama/paper.tex
Refined descriptions of formal proof, added more code excerpts

\paragraph{Ideals of commutative rings.}
The concept of an {ideal} of a ring can be thought of as an extension of the
notion of an integer $x$ in the ring of integers $\mathbb{Z}$. An \emph{ideal}
of commutative ring $R$ is a subset $I$ of $R$ that is itself an abelian group
under addition, which also satisfies the following ``absorption'' property:
Given any element $a$ of $I$, the product  $x \cdot a$ is again in $I$ for any
ring element $x \in R$.

@@

@@

In fact, $3 \mathbb{Z}$ is the only other maximal ideal of $\mathbb{Z}$
containing $6 \mathbb{Z}$, and in general, the prime ideals in the ring of
integers besides the zero ideal are those of the form $p \mathbb{Z}$, where $p$
a prime number.

@@

\paragraph{Module over commutative rings.}
Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a
binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted
$\cdot$, satisfying the following compatibility properties for all $r, s \in R$
and $u, v \in M$.

@@

\subsection{Building our Algebraic Hierarchy}
\begin{figure}[t]
  \caption{The hierarchy of our algebraic structures}
  \includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
  \centering
\end{figure}

@@

%While we found later on that we did not need quotients, 
While in the end, our formal proof did not call upon quotients of algebraic
structures, quotient rings and quotient modules are fundamental to commutative
algebra, and one can use them to construct alternate proofs of Nakayama's
Lemma. It is worth pointing out that we have formalized quotients of algebraic
objects in Coq using typeclasses, which appear to work rather nicely.

@@

The importance of quotients in commutative algebra motivates our use of
equivalence relations to define the components of a group structure. If one
were to instead use the regular Leibniz equality, it would be very difficult to
identify a quotient group with another group. However, by  defining a group in
terms of an arbitrary equivalence relation, we enable our theory to state that
a quotient group is simply a group under a different equivalence. Not much is
lost, as Coq's setoid rewrite tactics can still be called upon; a setoid is a
type equipped with an equivalence relation.

Moving onward, rings are the next step in our algebraic heirarchy; a ring has two binary
operations: addition, which must be commutative, and multiplication, which need
not be commutative in general. Next, we formalized the definition of a commutative
ring, further requiring commutativity of multiplication, as well as a
multiplicative identity. At this point, we formalized the notion of an ideal of
a commutative ring, a subgroup of the ring under addition that satisfies the
absorption property under multiplication, i.e., \(r a\) is in the ideal for
every element \(a\) of the ideal, and every element \(r\) of the commutative
ring. We also used this to formalize the notion of a quotient ring $R/I$, where
$I$ is an ideal of a commutative ring $R$.

Next, we formally defined a maximal ideal, a proper ideal that are maximal with
respect to inclusion, i.e., if it is strictly contained in another ideal, this
larger ideal must be the entire ring. Below is the definition in Coq, which
uses \texttt{P} as the predicate for the ideal.

@@

Using the definition of a maximal ideal, we then were able to define a local
ring, i.e., a commutative ring with a single maximal ideal.

@@

At this point we formalized the definition of a module over a commutative ring,
the commutative-algebraic generalization of the notion of a vector space over a
field. Nakayama's Lemma is a statement about finitely generated modules, and
hence we must formalize the notion of a scalar combination of a fixed finite
collection of elements $u_1, \ldots, u_n$ of an $R$-module $M$, i.e.,
expressions of the form $r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each
$r_i \in R$.

@@

\subsection{Constructing the Formal Proof}
Beyond formalizing the relevant structures from higher algebra, we also
formally establish some basic theory. For instance, a \emph{unit} of a
commutative ring $R$ is an element $x \in R$ with a multiplicative inverse,
i.e., an element $x^{-1} \in R$ for which $x \cdot x^{-1}$ is the
multiplicative identity $1 \in R$.

@@

    If \(I_{1}\) is not a maximal ideal, then there exists a strictly larger
    proper ideal \(I_{2}\) of $R$, i.e.,
    $x\in I_{1}\subsetneq I_{2} \subsetneq R$. Moreover, if  \(I_{2}\) is not
    maximal, then there exists a strictly larger proper ideal
    $I_{3} \supsetneq I_2$.

@@

    It is straight-forward to show that \(\bigcup_{k=1}^{\infty} I_{k}\)
    satisfies the definition of an ideal of $R$, and this union certainly
    contains \(x\). So by Zorn's lemma, since every ascending chain of ideals
    containing $x$ with respect to inclusion has its union as an upper bound,
    there exists a maximal ideal of \(R\) containing \(x\).
\end{quote}

This argument has potentially infinitely many steps, and chose to avoid this
issue by including an axiom that in any non-unit \(x\) is contained in some
maximal ideal.

\todo[inline]{Include some relevant code}

We also used classical logic to prove that \(1 - x\) is a unit where \(x\) is
a non-unit in any local ring. The proof used the rule \(\neg\neg P\rightarrow
P\) in order to do a proof by contradiction.

\todo[inline]{Include some relevant code}

Finally, we turn to our formal proof of Nakayama's Lemma, our ultimate goal.

@@

We needed a lemma before moving on to prove the main theorem which states that
for a finitely generated $R$-module \(M\) with generating set
\(b_{1}, b_{2}, \dots, b_{n}\), any element \(x\in I M\), where \(I\) is an
ideal of $R$, can be written as a linear combination of the generating set with
coefficients coming from \(I\).
\emily{I'm confused:  Isn't this how we defined $I M$? Where are the $u$'s and
  $a$'s coming from?}
In other words, given \(x\in I M\), there is by definition a linear combination
between elements \(u_{1}, \dots, u_{m}\in I\), and elements
\(a_{1},\dots, a_{m}\in M\), we need a lemma that states \(x\) can be written
as a linear combination between elements of \(I\) and the generating set
\(b_{1},\dots, b_{n}\in M\). This follows the informal mathematical argument
going inductively through the elements needed to generate \(x\) and shows that
each vector can be rewritten in terms of the generating set times coefficients
from the ideal \(I\) as follows.
\begin{align*}
  x & = \sum_{k=1}^{m} u_{k} a_{k} \\
    & = \sum_{k=1}^{m} u_{k} (r_{k1} b_{1} + \cdots + r_{kn} b_{n}) \\
    & = \sum_{j=1}^{n} (u_{1} r_{1j} + \cdots + u_{m} r_{mj}) b_{j}
\end{align*}
Since each \(u_{k}\) comes from the ideal \(I\), the absorbing property of
ideals guarantees that \(u_{k} r_{k j}\) is also contained in \(I\). As ideals
are also closed under addition, it follows that \(u_{1} r_{1j} + \cdots +
u_{m} r_{m j}\) are all elements of \(I\). Thus, \(x\) can be written as a
linear combination of the generating set with the coefficients coming from
\(I\).

\emily[inline]{I think we need to include more code here!  There are a lot of
  pieces of formal proof that would be illustrative.}

Now we move to describe the skeleton of our formal proof of Nakayama's lemma.
We proceed by induction on the number of elements in a generating set for the
finitely-generated module $M$. The base case where \(M\) is generated by 0
elements is by definition true since an empty linear combination is
conventionally taken to be the zero vector.
\emily{I think we should consider using one generator for the base case.}
The inductive case where \(M\) is generated by the vectors
\(b_{1}, \dots, b_{m}, b_{m+1}\). By assumption \(M = \mathfrak{m} M\).
With \(b_{1}\in M\), then \(b_{1}\in\mathfrak{m} M\) meaning that by our lemma,
there exists a linear combination for \(b_{1}\), say
\[b_{1} = u_{1} b_{1} + \cdots + u_{m+1} b_{m+1}\]
for some \(u_{1}, \dots, u_{m+1}\in\mathfrak{m}\). Collect the \(b_{1}\) terms
on the left-hand side of the equation to get that
\[(1 - u_{1}) b_{1} = u_{2} b_{2} + \cdots + u_{m+1} b_{m+1}\text{.}\]
As mentioned when defining maximal ideals, \(\mathfrak{m}\) can only contain
non-units. Namely, \(u_{1}\) must be a non-unit. Then we had already proven
that \(1 - u_{1}\) is a unit as we are in a local ring. Let \(v_{1}\) be the
multiplicative inverse of \(1 - u_{1}\), and multiply it on both sides of the
equation.
\begin{align*}
  v_{1} (1 - u_{1}) b_{1}
    & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
  1 b_{1} & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
  b_{1} & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
\end{align*}
We have thus found that one of the basis vectors is not needed to generate this
module. Using the induction hypothesis, we have that \(M = 0\). Showing that
the induction hypothesis holds in Coq takes more work than in an informal
proof, but this extra work is just a lot of bookkeeping.

>>>>

\paragraph{Ideals of commutative rings.}
The concept of an {ideal} of a ring can be thought of as an extension of the
notion of an integer $n$ in the ring of integers $\mathbb{Z}$. An \emph{ideal}
of commutative ring $R$ is a subset $I$ of $R$ that is itself an abelian group
under addition, which also satisfies the following ``absorption'' property:
Given any element $a$ of $I$, the product  $x \cdot a$ is again in $I$ for any
ring element $x \in R$.

@@

A \emph{prime ideal} of a commutative ring is a proper ideal $I$ with the following property:  
If the product $x \cdot y$ of ring elements $x$ and $y$ is in $I$, then $x \in I$ or $y \in I$.
The naming convention is motivated by the ring of integers, where the prime ideals are precisely those 
of the form $p\mathbb{Z}$, where $p$ is a prime number, along with the zero ideal.

@@

In fact, $3 \mathbb{Z}$ is the only other maximal ideal of $\mathbb{Z}$
containing $6 \mathbb{Z}$, and in general, the prime ideals in the ring of
integers besides the zero ideal are  $p \mathbb{Z}$, where $p$
a prime number. 
It is not a coincidence that every maximal ideal of the ring of integers is also a prime 
ideal; the analogous statement can be proved in arbitrary commutative rings.

@@

\paragraph{Modules over commutative rings.}
Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a
binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted
$\cdot$, satisfying the following properties for all $r, s \in R$
and $u, v \in M$.

@@

\subsection{Our Algebraic Hierarchy}
\begin{figure}[t]
  \caption{The hierarchy of our algebraic structures}
  \includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
  \centering
\end{figure}

@@

%While we found later on that we did not need quotients, 
While in the end, our formal proof
does not require calling upon 
 quotients of algebraic
structures, quotient rings and quotient modules are fundamental to commutative
algebra, and one can use them to construct alternate proofs of Nakayama's
Lemma. It is worth pointing out that we have formalized quotients of algebraic
objects in Coq using typeclasses, which appear to work rather nicely.

@@

The importance of quotients in commutative algebra motivates our use of
equivalence relations to define the components of a group structure. If one
were to instead use the regular Leibniz equality, it would be difficult to
identify a quotient group and another group. However, by  defining a group in
terms of an arbitrary equivalence relation, in our theory a quotient group is 
simply defined as a group, but under an equivalence relation that is not the usual 
equality. 
Not much is lost due to Coq's rewrite tactics for setoids--types equipped with an equivalence relation--
can still be called upon.

Moving onward, rings form the next step in our algebraic hierarchy; a ring has two binary
operations: addition, which must be commutative, and multiplication, which need
not be commutative in general. Next, we formalized the definition of a commutative
ring, further requiring commutativity of multiplication, as well as a
multiplicative identity. At this point, we formalized the notion of an ideal of
a commutative ring, a subgroup of the ring under addition that satisfies the
absorption property under multiplication, i.e., \(r a\) is in the ideal for
every element \(a\) of the ideal, and every element \(r\) of the commutative
ring. We also used this to formalize the notion of a quotient ring $R/I$, where
$I$ is an ideal of a commutative ring $R$.

Next, we formalized the definition of a  prime ideal, and then 
moved on to do the same for the notion of a maximal ideal, a proper ideal 
that is maximal with respect to inclusion. Below is the definition in Coq, which
uses \texttt{P} as the predicate for the ideal.

@@

We then were able to take advantage of the definition of a maximal ideal
to formally define a local
ring, i.e., a commutative ring with a single maximal ideal.

@@

At this point we formalized the definition of a module over a commutative ring,
the commutative-algebraic generalization of the notion of a vector space over a
field. Nakayama's Lemma is a statement about finitely generated modules, and
hence we must formalize the notion of a scalar combination of a finite
collection of elements, $u_1, \ldots, u_n$, of an $R$-module $M$, i.e.,
expressions of the form $r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each
$r_i \in R$.

@@

\subsection{Constructing the Formal Proof}
Beyond formalizing the relevant structures from higher algebra, we also
formally establish some basic theory. For instance, a \emph{unit} of a
commutative ring $R$ is an element $x \in R$ with a multiplicative inverse,
i.e., an element $x^{-1} \in R$ for which $x \cdot x^{-1}$ is the
multiplicative identity $1 \in R$.  In fact, it $x$ is a unit, it has a unique  inverse.

@@

    If \(I_{1}\) is not a maximal ideal, then there exists a strictly larger
    proper ideal \(I_{2}\) of $R$, i.e.,
    $x\in I_{1}\subsetneq I_{2} \subsetneq R$. Moreover, if  \(I_{2}\) is not
    maximal, then there exists a strictly larger proper ideal
    $I_{3}$ containing $I_2$.

@@

    It is straightforward to verify that the increasing union of all $I_{k}$, $k \geq 1$, 
    is by definition an ideal of $R$, and certainly
    contains \(x\).  Hence by Zorn's lemma, since every ascending chain of ideals
    containing $x$ with respect to inclusion has its union as an upper bound,
    there exists a maximal ideal of \(R\) containing \(x\).
\end{quote}

This argument has potentially infinitely many steps, and chose to avoid this
issue by including an axiom that in any non-unit \(x\) of a ring is contained in some
maximal ideal.

\begin{minted}{coq}
Axiom comm_ring_nonunit_maximal_ideal:
  forall (x: Carrier),
    ~ is_unit equiv mul one x ->
    exists (P: Carrier -> Prop)(P_proper: Proper (equiv ==> iff) P)
    	(P_ideal: Ideal add zero minus mul P),
      P x /\ maximal_ideal P.
\end{minted}

We also used classical logic to prove that \(1 - x\) is a unit whenever \(x\) is
an element of a local ring that is not a unit. The proof is completed by way of contradiction, and uses the rule that  
 \(\neg\neg P\rightarrow
P\).

To start describing our proof, we formalize the statement of Nakayama's Lemma in Coq.

@@

\begin{minted}{coq}
Theorem nakayama:
  forall {n: nat}(basis: t M n),
    finitely_generated Mequiv Madd Mzero action basis ->
  (forall a: M, ideal_module_pred a) ->
  forall a: M, a =M= Mzero.
\end{minted}

To use as a building block in the formal proof of Nakayama's Lemma, we first formally stated and proved a lemma that gives a concrete description of the elements of the submodule $\mathfrak{m} M$  appearing in the statement of the the theorem.  
This lemma applies more generally, to any finitely generated module $M$ over a (not necessarily local) commutative ring, and any submodule of the form $IM$, for $I$ an arbitrary ideal. 
By definition, $IM$ consists of scalar combinations of elements of $M$ with coefficients in $I$. 
The lemma states that the elements from $M$ appearing in the expression can be chosen to be from any 
 fixed finite generating set for $M$. 
In other words, given any generating set  $u_{1},\dots, u_{n} \in M$  of a  finitely generated $R$-module $M$, every element $x$ of $I M$ can be written as a scalar combination $a_1 \cdot u_1 + a_2  \cdot u_2 + \cdots + a_n  \cdot u_n$, where each $a_i \in I$.

\begin{minted}{coq}
Lemma module_fin_gen_ideal_module:
  forall {n: nat}(basis: t M n),
    finitely_generated Mequiv Madd Mzero action basis ->
    forall {m: nat}(coeffs: t R m)(vectors: t M m),
      Forall P coeffs ->
      exists (coeffs': t R n),
        linear_combin Madd Mzero action coeffs vectors =M=
          linear_combin Madd Mzero action coeffs' basis /\
        Forall P coeffs'.
\end{minted}

The proof follows from a straightforward argument based on definitions, by induction on the number of elements in a fixed generating set for $M$, which we informally describe:  By definition, $x \in I M$ can  be written, for some positive integer $k$ and elements $r_i \in R$ and $w_i \in M$,  $1 \leq i \leq k$, as   $x = r_1 \cdot w_1 + r_2 \cdot w_2 + \cdots + r_k \cdot w_k$.
What's more, by definition of a finite generating set $u_{1},\dots, u_{n}$ for $M$,  each $w_i$ equals 
$c_{i1} \cdot u_1 + c_{i2} \cdot u_2 + \cdots + c_{in} \cdot u_n$ for appropriate choices of $c_{ij} \in R$. 
Hence, inductively applying associativity, 
\begin{align*}
  x = \sum_{i=1}^{k} r_i  \cdot  w_i  =  \sum_{i=1}^{k} r_i  \cdot  \left(  \sum_{j=1}^n c_{ij}  \cdot  u_j \right) 
  =   \sum_{j=1}^n  \sum_{i=1}^{k} r_i   \cdot  (c_{ij}  \cdot  u_j ) 
=  \sum_{j=1}^n \left( \sum_{i=1}^{k}  c_{ij}  \cdot   r_i \right)   \cdot  u_j.   
\end{align*}
By absorption in ideals, each  $c_{ij}  \cdot r_i$ is in $I$, and since ideals are closed under addition, we inductively conclude that 
the coefficient $a_j := c_{1j} \cdot  r_1 +  c_{2j} \cdot  r_2 +  \cdots c_{kj}\cdot r_k$ of $u_j$ is also in $I$.

Now we move forward to outline the formal proof of Nakayama's lemma.
We proceed by induction on the number of elements in a generating set for the
$R$-module $M$. 
The base case is the situation when \(M\) requires no generators, so 
$M$ consists solely of the empty scalar combination, i.e., the empty sum, which, by convention, is the zero element.
In other words, $M = 0$ by assumption, and this is also precisely the 
conclusion of Nakayama's Lemma.  Hence the statement trivially, without using the hypothesis that $M = \mathfrak{m} M$.

We turn to the inductive step.  Fixing an arbitrary nonnegative integer $n$, 
we assume that Nakayama's Lemma holds in the case that $M$ has a generating set 
consisting of $n$ elements.

Now, 
fix an $R$-module $M$ generated by $u_1, \ldots, u_{n+1} \in M$. 
By assumption, \(M = \mathfrak{m} M\), and in particular, 
the generator $u_1$ is an element of the submodule $\mathfrak{m}M$.
The lemma described above guarantees the existence of ring elements $a_1, \ldots, a_n \in \mathfrak{m}$
for which 
\[u_{1} = a_{1} \cdot u_{1} + a_2 \cdot u_2 + \cdots + a_{n+1} \cdot u_{n+1}. \]

Collecting the \(u_{1}\) terms
on the left-hand side of the equation, applying the existence of additive inverses in $R$, and
 distributivity of scalar multiplication for modules, we see that
\begin{equation} \label{equation1: e}
(1 - a_{1}) \cdot u_{1} = a_{2} \cdot u_{2} + \cdots + a_{n+1} \cdot u_{n+1}\text{.}
\end{equation}

Since a proper ideal contains no units and $a_1$ is taken to be in $\mathfrak{m}$, it is not a unit.  
Hence, by the formalized statement mentioned earlier, $1-a_1$ is a unit of $R$.    
Let $b_{1} \in R$ denote its 
multiplicative inverse, so that $b_1 \cdot  (1 - a_{1}) = 1$.
Multiplying this element on the either side of \eqref{equation1: e}, we deduce the following.  
\begin{align*}
b_1 \cdot \left(  (1 - a_{1}) \cdot u_{1} \right)  & = b_1 \cdot  ( a_{2} \cdot u_{2} + \cdots + a_{n+1} \cdot u_{n+1} ) \\
\left( b_1 \cdot  (1 - a_{1})\right) \cdot u_{1}   & = b_1 \cdot  ( a_{2} \cdot u_{2} ) + \cdots + b_1  \cdot   (a_{n+1} \cdot u_{n+1} )  \\
u_1 = 1 \cdot u_{1}   & = ( b_1 \cdot  a_{2}  ) \cdot u_{2}  + \cdots + ( b_1  \cdot  a_{n+1} ) \cdot u_{n+1}  
\end{align*}
In particular, the generator $u_1$ can be written as a scalar combination of the $n$ generators $u_2, \ldots, u_{n+1}$, and is superfluous. Hence $M$ can be generated by $n$ elements, and $M=0$ by the inductive hypothesis.

Showing that
the inductive hypothesis holds in Coq is more work than in this informal
proof, but this extra work is just a lot of bookkeeping.
